---
title: 'Tutorial: distance calculations'
author: "EB, PJ, MG, CR"
date: "March 21, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting

This script compares different distances between data sets, in a multivariate, i.i.d. setting.

The model specifies $Y \sim \mathcal{N}(\mu, \Sigma)$, where $\Sigma$ is a fixed covariance matrix.
The prior on $\mu$ is $\mathcal{N}(0,1)$ on each component.

We begin by loading the package, registering multiple cores, setting the random number generator, etc.

```{r init, message=FALSE,warning=FALSE}
# load package
library(winference)
# register parallel cores
registerDoParallel(cores = detectCores())
# remove all
rm(list = ls())
# apply preferences for ggplotting
require(gridExtra)
theme_set(theme_bw())
# set RNG seed
set.seed(11)
```

## Data and model
We define a multivariate Gaussian model and generate some data.
```{r generatedata, tidy=T}
# number of observations
dimension <- 5
target <- get_multivariate_normal(dimension)
target$parameters$S <- diag(1, dimension, dimension)
for (i in 1:dimension){
  for (j in 1:dimension){
    target$parameters$S[i,j] <- 0.5^(abs(i-j))
  }
}
nobservations <- 100
target$simulate <- function(theta)
  target$robservation(nobservations, theta, target$parameters,  target$generate_randomness(nobservations))
# number of observations
true_theta <- rnorm(dimension)
obs <- target$simulate(true_theta)
# the observations are in a (ydim,nobservations) matrix
dim(obs)
```

## Distances

We consider four distances between multivariate samples: 

* the *exact Wasserstein* distance, using the transport package (Schuhmacher, D., Bhre, B., Gottschlich, C. and Heinemann, F. (2017) transport: Optimal Transport in Various Forms),

* the *Sinkhorn distance* (Cuturi, M. (2013) Sinkhorn distances: lightspeed computation of optimal transport. In Advances in Neural Information Processing Systems (NIPS), 2292–2300),

* the *Hilbert distance*, based on the Hilbert space filling curve, proposed in our article (Bernton, E., Jacob, P. E., Gerber, M. and Robert, C. P. (2017) Inference in generative models using the Wasserstein distance. arXiv preprint arXiv:1701.05146),

* and finally the *swapping distance*, proposed in (Puccetti, G. (2017) An algorithm to approximate the optimal expected inner product of two vectors with given marginals. Journal of Mathematical Analysis and Applications, 451, 132–145). 

Let's compute these distances between two data sets, playing a bit with the tuning parameters
of the Sinkhorn distance, that is, the regularizer $\varepsilon$ and the number of Sinkhorn iterations.

```{r test}
# generate a fake data set
fake_obs <- target$simulate(rnorm(dimension))
exact_transport_distance(obs, fake_obs, p = 1, ground_p = 2)
sinkhorn_distance(obs, fake_obs, p = 1, ground_p = 2, eps = 0.1, niterations = 100)$corrected
sinkhorn_distance(obs, fake_obs, p = 1, ground_p = 2, eps = 0.01, niterations = 100)$corrected
sinkhorn_distance(obs, fake_obs, p = 1, ground_p = 2, eps = 0.01, niterations = 1000)$corrected
swap_distance(obs, fake_obs, p = 1, ground_p = 2, tolerance = 1e-5)$distance
hilbert_distance(obs, fake_obs, p = 1, ground_p = 2)
```
We see that these distances take similar values: in fact the last three upper-bound the first. 

## Inference using the Hilbert distance

Now let's infer the parameters using the Hilbert distance. We specify the Hilbert distance
to the *wsmc* function, which runs an SMC sampler with r-hit MCMC rejuvenation steps.


```{r hilbertestimation, cache = T, message=F, warning=F, tidy=T}
param_algo <- list(nthetas = 1024, nmoves = 1, proposal = mixture_rmixmod(),
                   minimum_diversity = 0.5, R = 2, maxtrials = 1e5)
compute_distance <- get_hilbert_to_y(obs)
wsmcresults_hilbert <- wsmc(compute_distance, target, param_algo, maxtime = 20)
```

The *wsmc* function outputs the above information to monitor the progress of the sampler, 
as the threshold decreases. We can plot the resulting approximation of the posterior distribution, for instance as follows,
where the 2d-density plots obtained for the sequence of thresholds are overlaid. We focus on one of the bivariate marginals.

```{r plothilbert, dependson="hilbertestimation", tidy=T}
plot_bivariate_polygon(wsmcresults_hilbert, i1 = 3, i2 = 4) + geom_vline(xintercept = true_theta[3]) + geom_hline(yintercept = true_theta[4]) + xlab("X3") + ylab("X4")
```

We see the WABC posterior distributions concentrating around the data-generating values, indicated
by vertical and horizontal lines.

## Distance comparison

Based on the resulting samples, we can compare the distances to the exact Wasserstein distances.
To save calculations, we resort to the functions that compute the distances given a pre-computed 
cost matrix.

```{r comparedistances, cache = T, dependson="hilbertestimation", warning=F, message=F, tidy=T}
w1 <- rep(1/nobservations, nobservations)
w2 <- rep(1/nobservations, nobservations)
y_samples <- wsmcresults_hilbert$latest_y
d_comparison <- foreach(i = 1:length(y_samples), .combine = rbind) %dorng% {
  C <- cost_matrix_L2(obs, y_samples[[i]])
  hilbert <- hilbert_distance(obs, y_samples[[i]], p = 1, ground_p = 2)
  exact <- as.numeric(exact_transport_given_C(w1, w2, C, p = 1))
  sinkhorn1 <- sinkhorn_given_C(w1, w2, C, p = 1, eps = 0.05, niterations = 100)$corrected
  sinkhorn2 <- sinkhorn_given_C(w1, w2, C, p = 1, eps = 0.025, niterations = 1000)$corrected
  swap <- swap_distance(obs, y_samples[[i]], p = 1, ground_p = 2, tolerance = 1e-5)$distance
  data.frame(hilbert = hilbert,
             exact = exact,
             swap = swap,
             sinkhorn1 = sinkhorn1,
             sinkhorn2 = sinkhorn2)
}
g <- qplot(x = d_comparison$exact, y = d_comparison$hilbert, geom = "blank")
g <- g + geom_point(aes(colour = "hilbert")) + geom_abline(slope = 1, intercept = 0)
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$sinkhorn1, colour = "sinkhorn 1"))
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$sinkhorn2, colour = "sinkhorn 2"))
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$swap, colour = "swap"))
g <- g + xlab("exact Wasserstein") + ylab("approximation") + scale_colour_manual(name = "", values = c("black", "orange", "blue", "red"))
g 
```

We see that the approximate distances (Hilbert, Sinkhorn, swapping) are all above the exact Wasserstein distances.
Furthermore, they seem to be pretty close to the exact distances.
However, if we pursue the inference by doing more SMC steps, 
the parameters concentrate, and then we see more difference between the distances.

```{r continued, cache = T, dependson="hilbertestimation", warning=F, message=F, tidy=T}
wsmcresults_hilbert_continued <- wsmc_continue(wsmcresults_hilbert, maxstep = 5)
y_samples <- wsmcresults_hilbert_continued$latest_y
d_comparison <- foreach(i = 1:length(y_samples), .combine = rbind) %dorng% {
  C <- cost_matrix_L2(obs, y_samples[[i]])
  hilbert <- hilbert_distance(obs, y_samples[[i]], p = 1, ground_p = 2)
  exact <- as.numeric(exact_transport_given_C(w1, w2, C, p = 1))
  sinkhorn1 <- sinkhorn_given_C(w1, w2, C, p = 1, eps = 0.05, niterations = 100)$corrected
  sinkhorn2 <- sinkhorn_given_C(w1, w2, C, p = 1, eps = 0.025, niterations = 1000)$corrected
  swap <- swap_distance(obs, y_samples[[i]], p = 1, ground_p = 2, tolerance = 1e-5)$distance
  data.frame(hilbert = hilbert,
             exact = exact,
             swap = swap,
             sinkhorn1 = sinkhorn1,
             sinkhorn2 = sinkhorn2)
}
g <- qplot(x = d_comparison$exact, y = d_comparison$hilbert, geom = "blank")
g <- g + geom_point(aes(colour = "hilbert")) + geom_abline(slope = 1, intercept = 0)
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$sinkhorn1, colour = "sinkhorn 1"))
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$sinkhorn2, colour = "sinkhorn 2"))
g <- g + geom_point(aes(x = d_comparison$exact, y = d_comparison$swap, colour = "swap"))
g <- g + xlab("exact Wasserstein") + ylab("approximation") + scale_colour_manual(name = "", values = c("black", "orange", "blue", "red"))
g 
```

We see that the difference between the Hilbert distance and the exact Wasserstein distances widens. However,
the Sinkhorn distance with a small value of $\epsilon$, and the swap distance, are still very close to the Wasserstein distances.
